---
title: "Assignment 2: Coding Basics"
author: "Griffin Bird"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

```{r 1 to 100 increasing by 4}
seq_by4 <- seq(0, 100, by = 4)

head(seq_by4)
```

2.  Compute the mean and median of this sequence.

```{r}
seq_mean <- mean(seq_by4)
seq_median <- median(seq_by4)
```

3.  Ask R to determine whether the mean is greater than the median.

```{r}
seq_mean > seq_median
```

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. I used the sequence function to generate numbers from 1 to 100, and used "by" to indicate that the sequence should increase by increments of 4

#2. I calculated the mean and median of the data set using Rs built in functions, then assigned those values to variables

#3. I used a boolean expression to get R to tell me whether the mean was greater than the median, R will return true or false, and it turns out the mean is not greater than the median.

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

```{r}

students <- c('John', 'Jane', 'Jill', 'Joseph', 'Jacqueline', 'Josephine')
#Character vectors

scores <- runif(6, min=0, max=100)
#Numeric vectors

pass <- (scores>50)
#Logical vectors

```


7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

```{r}

tests.df <- data.frame(pass, scores, students)

```


8.  Label the columns of your data frame with informative titles.

```{r}
colnames(tests.df) <- c('Passed', 'Test Score', 'Student Name')
```


9.  QUESTION: How is this data frame different from a matrix?

> Answer:
The data frame contains different types of vectors, a matrix can only contain one type of vector

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

```{r}

did.they.pass <- function(x) { ifelse(x >50, "They passed", "They failed")}

```


11. Apply your function to the vector with test scores that you created in number 5.

```{r}

did.they.pass(scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
