---
title: "Assignment 3: Data Exploration"
author: "Griffin Bird"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code.

------------------------------------------------------------------------

## Set up your R session

1.  Check your working directory, load necessary packages (tidyverse, lubridate), and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
setwd("C:/Users/17038/Documents/EDA Spring 2023/")
getwd()

library(tidyverse)
library(lubridate)

Neonics <- read.csv("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv", stringsAsFactors = TRUE)  
Litter <- read.csv("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv", stringsAsFactors = TRUE)  

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Decreasing populations of pollinators is a pressing problem, and it's been established that insecticeds contibute to excess pollinator death. Makes sense that we would want to look the one of the most widely used classes of insecticides and their toxological impact on insects.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Forest litter influences the temperature, duration, and veracity of wildfires, so studying forest litter would give us a better idea of how a wildfire will act in a given area. Seeing how wildfires have plagued western states recently, I can imagine folks in CO are interested in what will fuel the next fire.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 1. The locations of tower plots, where woody debris sampling takes place, is randomly selected. 2. Trap placement within plots is either targeted or randomized, depending on vegetation. In sites with more than 50% aerial cover of woody vegetation greater than 2m in height, trap placement is randomized. In sites with less than 50% cover of woody vegetation or heterogenously distributed, patchy vegetation, trap placement is targeted.\
> 3. Ground traps are sampled once per year, and elevated traps are sampled at different frequencies depending on what vegetation is present.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(Neonics)
```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest?

```{r}
summary(Neonics$Effect)
```

> Answer: The two most commonly studied effects are Population and Mortality. These effects would be of particular interest because pesticide is known to harm insect populations, particularly pollinator populations. I imagine scientists are trying to figure out exactly how pollinators are affected by this class of pesticides.

7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: The `sort()` command can sort the output of the summary command...]

```{r}

summary(Neonics$Species.Common.Name) %>%
  sort(decreasing = TRUE)

```

> Answer: The six most commonly studied species, by common name, are: Honey Bee, Parasitic Wasp, Buff Tailed Bumblebe, Carniolan Honey Bee, Bumble Bee, Italian Honeybee. As it would happen, all of these species are pollinators! These are probably the most frequently studied for that exact reason, their populations are declining and we want to know what role insecticides play in that.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric?

```{r}

class(Neonics$Conc.1..Author.)

```

> Answer: `Conc.1..Author.` is factor data. R imported this column as factors because the column contains some non-numeric values, a lot of "/"s. R is kicking it all into factors because of those non-numeric characters.

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r, fig.height = 6, fig.width = 8}


ggplot(Neonics) + 
 geom_freqpoly(aes(x = Publication.Year), bins = 50) + ggtitle("Number of Publications by Year") + xlab("Publication Year")+ ylab("Count")

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r, fig.height = 6, fig.width = 8}

ggplot(Neonics) + 
 geom_freqpoly(aes(x = Publication.Year, color = Test.Location), bins = 50) + ggtitle("Number of Publications by Year") + xlab("Publication Year")+ ylab("Count") + labs(color = "Test Location") 

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are the lab and field natural. Field natural and lab experiments frequency are close through 2008 or so, but come 2015 lab is much more frequent than field natural.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information.

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r, fig.height = 6, fig.width = 8}


ggplot(Neonics) + 
 geom_bar(aes(Endpoint)) +  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + ylab("Count") + ggtitle("Endpoint Frequency")

```

> Answer: The two most common endpoints are LOEL and NOEL. Loel is the lowest observable effect level at which a pesticide produced effects significantly different than the response to controls. Noel is the no-observable-effect-level which is the highest dose producing effects not significantly different from responses of controls.

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}

class(Litter$collectDate)
```

```{r}
Litter$collectDate <- ymd(Litter$collectDate)
class(Litter$collectDate)
```

```{r}
unique(Litter$collectDate)
```

13. Using the `unique` function, determine how many plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

length(unique(Litter$plotID))

unique(Litter$plotID)
summary(Litter$plotID)

```

> Answer: 12 plots were sampled at Niwot Ridge. "Unique" generates a string of all unique values in a vector, "Summary" will do that and show a frequency count for each unique value.

14. Create a bar graph of functional Group counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r, fig.height = 5, fig.width = 10}

ggplot(Litter) + 
 geom_bar(aes(functionalGroup)) + xlab("Functional Group") + ylab("Count") + ggtitle("Functional Group Frequency")

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r, fig.height = 6, fig.width = 10}

ggplot(Litter)+
  geom_boxplot(aes(functionalGroup, dryMass)) + xlab("Functional Group") + ylab("Dry Mass") + ggtitle("Dry Mass by Functional Group")
```

```{r, fig.height = 6, fig.width = 10}
ggplot(Litter)+
  geom_violin(aes(functionalGroup, dryMass)) + xlab("Functional Group") + ylab("Dry Mass") + ggtitle("Dry Mass by Functional Group")
```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: So many funcitonal groups are being compared that it's difficult to see the definition of the "violin" for any one group, the box and whisker plot is much easier to read.

What type(s) of litter tend to have the highest biomass at these sites?

> Answer: Needles, twigs/branches, and mixed litter have the highest biomass.
