---
output:
  pdf_document: default
  html_document: default
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models.

## Directions

1.  Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up

1.  Set up your session:

-   Check your working directory
-   Load the tidyverse, lubridate, zoo, and trend packages
-   Set your ggplot theme

```{r data setup, message = FALSE}
#1
library(knitr)
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
library(ggplot2)
library(ggthemes)
library(zoo)
library(trend)
library(Kendall)

here
getwd()

mytheme <- theme_wsj() + theme(plot.title = element_text(hjust = 0.5), 
   panel.grid.minor = element_line(color = 2,
   size = 0.25,
   linetype = 1),
  legend.position = "bottom", 
  axis.text = element_text(face="plain", size = 12),
  axis.title=element_text(size=14),
  title = element_text(size=14))

theme_set(mytheme)

```

```{r data import}
#2

Garinger <- list.files(path="~/EDASpring2023/Data/Raw/Ozone_TimeSeries", full.names = TRUE) %>% 
  lapply(read_csv) %>% 
  bind_rows 

```
